An Intrinsically-Typed Solution for the\\ List-Machine Benchmark

### 2021-07-23 papers/COLA/paper.tex
Before revision

\begin{abstract}
Formal models are important tools in the programming language research
community. However, such models are full of intricacies and, due to that,
they are subject to subtle errors. Such failures motivated the usage of
tools to ensure the correctness of these formalisms. One way to eliminate
such errors is to encode models in a dependently-typed language in order
to ensure its ``correctness-by-construction''. In this paper, we use this
idea to build a verified interpreter for the list-machine benchmark in the
Agda programming language, comparing the results with formalizations developed
by Appel and Leroy. We formalize the 14 tasks of the benchmark
using roughly 14\% of LOC compared to a Twelf solution, and 47\% of LOC
compared to a Coq solution, even without the use of proof automation. We also
describe a solution to the second version of the benchmark and compare it with
Appel's and Leroy Coq-based solution.
\end{abstract}

@@

A popular approach for specifying formal semantics is the
use of definitional interpreters, which represents the meaning of a
programming language as an interpreter written in some
meta-language~\cite{Reynolds72}. A major advantage of such approach
is the possibility to validate the semantics through execution.
Recently, definitional interpreters were used for formalizing type
soundness theorems for some advanced typing features~\cite{Amin17},
and the semantics of imperative programming languages, in such a way the static
semantics is ensured by a dependently-typed
syntax\footnote{Also known as intrinsically-typed.}\cite{Poulsen18}.
In this work, we follow Poulsen et. al. by using an intrinsically-typed
representation to build a definitional interpreter for a low-level virtual
machine, developed by Appel and Leroy, as a benchmark problem closer to
real-world implementations such as typed assembly languages~\cite{CraryM99} and
proof-carrying code~\cite{Necula97}.

@@

More specifically, we contribute:

@@

Constructor \ensuremath{\Con{\lbrack\:\rbrack}} builds empty vectors. The cons-operator (\ensuremath{\Con{\_::\_}})
inserts a new element in front of a vector of $n$ elements (of type
\ensuremath{\D{Vec}\;\V{A}\;\V{n}}) and returns a value of type \ensuremath{\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})}. The
\ensuremath{\D{Vec}} datatype is an example of a dependent type, i.e., a type
that uses a value (that denotes its length). The usefulness of
dependent types can be illustrated with the definition of a safe list
head function: \ensuremath{\F{head}} can be defined to accept only non-empty
vectors, i.e.,~values of type \ensuremath{\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})}.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\F{head}\;\mathbin{:}\;\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})\;\to \;\V{A}{}\<[E]%
\\
\>[3]{}\F{head}\;(\V{x}\;\Con{::}\;\V{xs})\;\mathrel{=}\;\V{x}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

This type is called propositional equality. It defines that there is
a unique evidence for equality, constructor \ensuremath{\Con{refl}} (for reflexivity),
which asserts that the only value equal to \ensuremath{\V{x}} is itself. Given a predicate \ensuremath{\V{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set}}
and a vector \ensuremath{\V{xs}}, the type \ensuremath{\D{All}\;\V{P}\;\V{xs}} is used to build proofs that \ensuremath{\V{P}} holds for all
elements in \ensuremath{\V{xs}} and it is defined as:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{41}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathkw{data}\;\D{All}\;(\V{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set})\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\;\to \;{}\<[41]%
\>[41]{}\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{\lbrack\:\rbrack}\;\mathbin{:}\;\D{All}\;\V{P}\;\Con{\lbrack\:\rbrack}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{\_::\_}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{x}\;\V{xs}\mskip1.5mu\}\;\to \;\V{P}\;\V{x}\;\to \;\D{All}\;\V{P}\;\V{xs}\;\to \;\D{All}\;\V{P}\;(\V{x}\;\Con{::}\;\V{xs}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The first constructor specifies that \ensuremath{\D{All}\;\V{P}} holds for the empty vector and
constructor \ensuremath{\Con{\_::\_}} builds a proof of \ensuremath{\D{All}\;\V{P}\;(\V{x}\;\Con{::}\;\V{xs})} from proofs of
\ensuremath{\V{P}\;\V{x}} and \ensuremath{\D{All}\;\V{P}\;\V{xs}}. Since this type has the same structure of vectors,
some functions on \ensuremath{\D{Vec}} have similar definitions for type \ensuremath{\D{All}}. As an example
used in our formalization, consider the function \ensuremath{\F{lookup}}, which extracts a
proof of \ensuremath{\V{P}} for the element at position \ensuremath{\V{v}\;\mathbin{:}\;\D{Fin}\;\V{n}} in a \ensuremath{\D{Vec}}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[4]{}\F{lookup}\;\mathbin{:}\;\{\mskip1.5mu \V{xs}\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{All}\;\V{P}\;\V{xs}\;\to \;\V{P}\;\V{x}{}\<[E]%
\\
\>[4]{}\F{lookup}\;\Con{zero}\;(\V{px}\;\Con{::}\;\anonymous )\;\mathrel{=}\;\V{px}{}\<[E]%
\\
\>[4]{}\F{lookup}\;(\Con{suc}\;\V{idx})\;(\anonymous \;\Con{::}\;\V{pxs})\;\mathrel{=}\;\F{lookup}\;\V{idx}\;\V{pxs}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

An important application of dependent types is to encode programming languages
syntax. The role of dependent types in this domain is to encode programs that
only allow well-typed and well-scoped terms~\cite{Benton2012}. Intuitively, we encode
the static semantics of the object language in the host language AST's
constructor, leaving the responsibility of checking type safety to the
host's language type-checker. As an example, consider the following simple
expression language. For this paper, we assume a basic knowledge of functional
programming and Agda. %\footnote{For further information about Agda, see~\cite{Norell2009,Stump16}.}.

@@

With this data type\footnote{Agda supports the definition of mixfix operators.
We can use underscores to mark arguments positions.}, we can construct expressions
to denote terms that should not be considered well-typed like
\ensuremath{(\Con{Num}\;\V{1})\;\F{+}\;\Con{True}}. Using this approach, we need to specify the static semantics
as another definition, which should consider all possible cases to avoid the
definition of ill-typed terms.

@@

%Dependently typed pattern matching is built by using the so-called
%|with| construct, that allows for matching intermediate
%values~\cite{McBride2004}. If the matched value has a dependent type,
%then its result can affect the form of other values. For example,
%consider the following code that defines a type for natural number
%parity. If the natural number is even, it can be represented as the
%sum of two equal natural numbers; if it is odd, it is equal to one
%plus the sum of two equal values. Pattern matching on a value of
%|Parity n| allows to discover if $n = j + j$ or $n = S (k + k)$,
%for some $j$ and $k$ in each branch of |with|.  Note that the
%value of $n$ is specialized accordingly, using information ``learned''
%by the type-checker.
%\begin{spec}
%data Parity : Nat -> Set where
%   Even : forall {n : Nat} -> Parity (n + n)
%   Odd  : forall {n : Nat} -> Parity (S (n + n))
%
%parity : (n : Nat) -> Parity n
%parity = -- definition omitted

%natToBin : Nat -> List Bool
%natToBin zero = []
%natToBin k with (parity k)
%   natToBin (j + j)     | Even = false :: natToBin j
%   natToBin (succ (j + j)) | Odd  = true  :: natToBin j
%\end{spec}

@@

The list-machine, proposed by Appel et. al.~\cite{AppelDL12}, is a simple pointer virtual machine where values
are empty lists and cons-cells:
\[
a ::= nil\,\mid\,cons(a_1,a_2)
\]
Throughout this text, the meta-variable $a$ is used to denote an arbitrary value, $v$ denotes a variable
and $l$ a program label. Following common practice, all meta-variables can
appear primed or subscripted. The syntax of the virtual machine instructions are
presented next and their meaning is as usual.
\[
\begin{array}{rcll}
  \iota & ::=  & \text{jump }l                       & \text{(jump instruction)}\\
        & \mid & \text{branch-if-nil $v$~$l$}            & \text{(if $v = nil$ goto $l$)}\\
        & \mid & \text{fetch-field $v$ 0 $v'$}       & \text{(fetch the head of $v$ into $v'$)}\\
        & \mid & \text{fetch-field  $v$ 1 $v'$}      & \text{(fetch the tail of $v$ into $v'$)}\\
        & \mid & \text{cons $v_0$ $v_1$ $v'$}        & \text{(make a cons cell in $v'$)} \\
        & \mid & \text{halt}                         & \text{(finishes execution)}\\
        & \mid & \iota_1;\iota_2                             & \text{(sequential composition)}\\
      p & ::=  & l_i \,:\,\iota\,;\,p                    & \text{(program: sequence of blocks)}\\
        & \mid & \text{end}                          & \text{(end of block list)}\\
\end{array}
\]
A program is just a sequence of blocks referenced by a unique label.

Each program variable is assigned to a list type, which is used to guarantee the safety when executing
fetch-field instructions that demands non-empty list arguments. In order to express such refinements, types are subject to a
subtyping relation. The meta-variable $\tau$ denotes an arbitrary type.
\[
\begin{array}{rcll}
  \tau & ::=  & \text{nil} & \text{(type for empty lists)}\\
       & \mid & \text{list }\tau & \text{(lists whose elements have type $\tau$)}\\
       & \mid & \text{listcons }\tau & \text{(non-empty lists of $\tau$)}\\
\end{array}
\]
The notation $\tau \subset \tau'$
denotes the subtyping judgment, which is defined as follows.
\[
\begin{array}{ccc}
  \inference{}
            {\tau\subset \tau}
            [refl]
  &
  \inference{}
            {nil \subset \tau}
            [nil]
  &
  \inference{\tau \subset \tau'}
            {list\:\tau\subset list\:\tau'}
            [list]\\\\
  \multicolumn{3}{c}{
  \inference{\tau \subset \tau'}
            {listcons\:\tau\subset list\:\tau'}
            [listcons]} \\\\
            \multicolumn{3}{c}{
            \inference{\tau \subset \tau'}
            {listcons\:\tau\subset listcons\:\tau'}
            [mixed]} \\\\
\end{array}
\]
Basically, the subtyping relation specifies that $nil$ (empty list type) is
subtype of any list type and $listcons\:\tau$ is a subtype of the $list\:\tau$.
The other rules specify that type constructors $list$ and $listcons$ respect
the subtyping relation. The list common supertype $\tau = \tau_1 \sqcap \tau_2$ of
$\tau_1$ and $\tau_2$ is defined as the smallest type such that $\tau_1$ and $\tau_2$
are subtypes of $\tau$.

Following the common practice, the meta-variable $\Gamma$ denotes an
environment binding variable names to their types. Notation $\{\}$ denotes an empty environment, $v : \tau , \Gamma$
denotes the operation of including a new entry for variable $v$ with type $\tau$
in $\Gamma$ and $\Gamma [v := \tau]$ denotes the environment which is identical to $\Gamma$, except
by the entry which associates variable $v$ with type $\tau$.
Subtyping is extended to contexts as follows.
\[
\begin{array}{cc}
  \inference{}
            {\Gamma \subset_{env} \{\}}[b1]
  &
  \inference{\tau' \subset \tau & \Gamma' \subset_{env} \Gamma_2}
            {v : \tau' , \Gamma' \subset_{env} v : \tau , \Gamma_2}
            [b2]
\end{array}
\]
The variable $\Pi$ is used to denote \emph{program typings}, i.e. finite mappings between
labels and typing contexts $\Gamma$, where notation $\Pi(l) = \Gamma$ denotes that
$\Gamma$ stores the types of variables on the entry point of the block labeled by $l$.
Using the previously defined notations, the typing rules for the list-machine are defined
as a syntax-directed judgment $\Pi \vdash_{\text{instr}} \Gamma \{ \iota \} \Gamma'$,
which intuitively means that the instruction $\iota$ transforms an input typing environment
$\Gamma$ into an output environment $\Gamma'$ under a fixed program typing $\Pi$. The typing
rules for the list-machine instructions are defined as follows.

>>>>

\begin{abstract}
Formal models are important tools in the programming language research
community. However, such models are full of intricacies and, due to that,
they are subject to subtle errors. Such failures motivated the usage of
tools to ensure the correctness of these formalisms. One way to eliminate
such errors is to encode models in a dependently-typed language in order
to ensure its ``correctness-by-construction''. In this paper, we use this
idea to build a verified interpreter for the list-machine benchmark in the
Agda programming language, comparing the results with formalizations developed
by Appel et. al. We formalize the 14 tasks of the benchmark
using roughly 14\% of LOC compared to a Twelf solution, and 47\% of LOC
compared to a Coq solution, even without the use of proof automation. We also
describe a solution to the second version of the benchmark and compare it with
Appel'et. al. Coq-based solution.
\end{abstract}

@@

A popular approach for specifying formal semantics is the
use of definitional interpreters, which represents the meaning of a
programming language as an interpreter written in some
meta-language~\cite{Reynolds72}. A major advantage of such approach
is the possibility to validate the semantics through execution.
Recently, definitional interpreters were used for formalizing type
soundness theorems for some advanced typing features~\cite{Amin17},
and the semantics of imperative programming languages, in such a way the static
semantics is ensured by a dependently-typed
syntax\footnote{Also known as intrinsically-typed.}\cite{Poulsen18}.
In this work, we follow Poulsen et. al. by using an intrinsically-typed
representation to build a definitional interpreter for a low-level virtual
machine, developed by Appel et. al., as a benchmark problem closer to
real-world implementations such as typed assembly languages~\cite{CraryM99} and
proof-carrying code~\cite{Necula97}.

@@

Let us summarize our contributions. %More specifically, we contribute:

@@

Constructor \ensuremath{\Con{\lbrack\:\rbrack}} builds empty vectors. The cons-operator (\ensuremath{\Con{\_::\_}})\footnote{Agda supports the definition of mixfix operators.
We can use underscores to mark arguments positions.}
inserts a new element in front of a vector of $n$ elements (of type
\ensuremath{\D{Vec}\;\V{A}\;\V{n}}) and returns a value of type \ensuremath{\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})}. The
\ensuremath{\D{Vec}} datatype is an example of a dependent type, i.e., a type
that uses a value (that denotes its length). The usefulness of
dependent types can be illustrated with the definition of a safe list
head function: \ensuremath{\F{head}} can be defined to accept only non-empty
vectors, i.e.,~values of type \ensuremath{\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})}.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\F{head}\;\mathbin{:}\;\D{Vec}\;\V{A}\;(\Con{suc}\;\V{n})\;\to \;\V{A}{}\<[E]%
\\
\>[3]{}\F{head}\;(\V{x}\;\Con{::}\;\V{xs})\;\mathrel{=}\;\V{x}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

This type is called propositional equality. It defines that there is
a unique evidence for equality, constructor \ensuremath{\Con{refl}} (for reflexivity),
which asserts that the only value equal to \ensuremath{\V{x}} is itself.

Dependently typed pattern matching is built by using the so-called
\ensuremath{\mathkw{with}} construct, that allows for matching intermediate
values~\cite{McBride2004}. If the matched value has a dependent type,
then its result can affect the form of other values. For example,
consider the following code that defines a type for natural number
parity. If the natural number is even, it can be represented as the
sum of two equal natural numbers; if it is odd, it is equal to one
plus the sum of two equal values. Pattern matching on a value of
\ensuremath{\D{Parity}\;\V{n}} allows to discover if $n = j + j$ or $n = S (k + k)$,
for some $j$ and $k$ in each branch of \ensuremath{\mathkw{with}}.  Note that the
value of $n$ is specialized accordingly, using information ``learned''
by the type-checker.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
\column{43}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathkw{data}\;\D{Parity}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\Con{Even}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\;\mathbin{:}\;\D{\mathbb{N}}\mskip1.5mu\}\;\to \;\D{Parity}\;(\V{n}\;\F{+}\;\V{n}){}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\Con{Odd}\;{}\<[9]%
\>[9]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\;\mathbin{:}\;\D{\mathbb{N}}\mskip1.5mu\}\;\to \;\D{Parity}\;(\V{S}\;(\V{n}\;\F{+}\;\V{n})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\F{parity}\;\mathbin{:}\;(\V{n}\;\mathbin{:}\;\D{\mathbb{N}})\;\to \;\D{Parity}\;\V{n}{}\<[E]%
\\
\>[B]{}\F{parity}\;\mathrel{=}\orange{\texttt{-- definition omitted}}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\F{natToBin}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{List}\;\D{Bool}{}\<[E]%
\\
\>[B]{}\F{natToBin}\;\Con{zero}\;\mathrel{=}\;\Con{\lbrack\:\rbrack}{}\<[E]%
\\
\>[B]{}\F{natToBin}\;\V{k}\;\mathkw{with}\;(\F{parity}\;\V{k}){}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\F{natToBin}\;(\V{j}\;\F{+}\;\V{j})\;{}\<[25]%
\>[25]{}\mid \;\Con{Even}\;\mathrel{=}\;\Con{false}\;\Con{::}\;\F{natToBin}\;\V{j}{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\F{natToBin}\;(\Con{suc}\;(\V{j}\;\F{+}\;\V{j}))\;\mid \;\Con{Odd}\;{}\<[35]%
\>[35]{}\mathrel{=}\;\Con{true}\;{}\<[43]%
\>[43]{}\Con{::}\;\F{natToBin}\;\V{j}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
A important application of Agda's dependently-typed pattern matching is
to rewrite a equality. As an example, consider the following code which
postulates the commutativity of natural number addition and a predicate
over natural numbers, \ensuremath{\D{P}}.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathkw{postulate}\;\F{plus\textrm{-}comm}\;\mathbin{:}\;(\V{a}\;\V{b}\;\mathbin{:}\;\D{\mathbb{N}})\;\V{→}\;\V{a}\;\F{+}\;\V{b}\;\D{\equiv}\;\V{b}\;\F{+}\;\V{a}{}\<[E]%
\\
\>[B]{}\mathkw{postulate}\;\D{P}\;\mathbin{:}\;\D{\mathbb{N}}\;\V{→}\;\D{Set}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\F{foo}\;\mathbin{:}\;(\V{a}\;\V{b}\;\mathbin{:}\;\D{\mathbb{N}})\;\V{→}\;\D{P}\;(\V{a}\;\F{+}\;\V{b})\;\V{→}\;\D{P}\;(\V{b}\;\F{+}\;\V{a}){}\<[E]%
\\
\>[B]{}\F{foo}\;\V{a}\;\V{b}\;\V{t}\;\mathkw{with}\;{}\<[18]%
\>[18]{}\V{a}\;\F{+}\;\V{b}\;{}\<[25]%
\>[25]{}\mid \;\F{plus\textrm{-}comm}\;\V{a}\;\V{b}{}\<[E]%
\\
\>[B]{}\F{foo}\;\V{a}\;\V{b}\;\V{t}\;{}\<[14]%
\>[14]{}\mid \;.\;(\V{b}\;\F{+}\;\V{a})\;\mid \;\Con{refl}\;\mathrel{=}\;\V{t}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Function \ensuremath{\F{foo}} uses the \ensuremath{\mathkw{with}} construct to pattern-match on the equality produced
by \ensuremath{\F{plus\textrm{-}comm}},  \ensuremath{\V{a}\;\F{+}\;\V{b}\;\D{\equiv}\;\V{b}\;\F{+}\;\V{a}}. Since the propositional equality type has only one
constructor, \ensuremath{\Con{refl}}, matching over it makes Agda type checker to replace
all occurrences of \ensuremath{\V{a}\;\F{+}\;\V{b}} to \ensuremath{\V{b}\;\F{+}\;\V{a}} in the current equation context. We can notice
this by the use of the \empth{dotted-pattern} \ensuremath{.\;(\V{b}\;\F{+}\;\V{a})} which indicates that the
only way that this equation is correct is when the first result of the \ensuremath{\mathkw{with}} construct
is \ensuremath{\V{b}\;\F{+}\;\V{a}}. The use of dependent pattern matching to rewrite equalities in functions is
so common in Agda that the language provides a syntax sugar that simplifies such structure.
The \ensuremath{\mathkw{rewrite}} construct allows the substitution of an equality in all values in a certain
equation. The \ensuremath{\F{foo}} function could be represented more concisely, using \ensuremath{\mathkw{rewrite}}, as:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\F{foo}\;\mathbin{:}\;(\V{a}\;\V{b}\;\mathbin{:}\;\D{\mathbb{N}})\;\V{→}\;\D{P}\;(\V{a}\;\F{+}\;\V{b})\;\V{→}\;\D{P}\;(\V{b}\;\F{+}\;\V{a}){}\<[E]%
\\
\>[B]{}\F{foo}\;\V{a}\;\V{b}\;\V{t}\;\mathkw{rewrite}\;\F{plus\textrm{-}comm}\;\V{a}\;\V{b}\;\mathrel{=}\;\V{t}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Another inductive predicate used in our formalization is the type \ensuremath{\D{All}\;\D{P}\;\V{xs}} which
ensures that all elements of a vector \ensuremath{\V{xs}} satisfy a predicate \ensuremath{\D{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set}}.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{41}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathkw{data}\;\D{All}\;(\D{P}\;\mathbin{:}\;\V{A}\;\to \;\D{Set})\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\;\to \;{}\<[41]%
\>[41]{}\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{\lbrack\:\rbrack}\;\mathbin{:}\;\D{All}\;\D{P}\;\Con{\lbrack\:\rbrack}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{\_::\_}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{x}\;\V{xs}\mskip1.5mu\}\;\to \;\D{P}\;\V{x}\;\to \;\D{All}\;\D{P}\;\V{xs}\;\to \;\D{All}\;\D{P}\;(\V{x}\;\Con{::}\;\V{xs}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The first constructor specifies that \ensuremath{\D{All}\;\D{P}} holds for the empty vector and
constructor \ensuremath{\Con{\_::\_}} builds a proof of \ensuremath{\D{All}\;\D{P}\;(\V{x}\;\Con{::}\;\V{xs})} from proofs of
\ensuremath{\D{P}\;\V{x}} and \ensuremath{\D{All}\;\D{P}\;\V{xs}}. Since this type has the same structure of vectors,
some functions on \ensuremath{\D{Vec}} have similar definitions for type \ensuremath{\D{All}}. As an example
used in our formalization, consider the function \ensuremath{\F{lookup}}, which extracts a
proof of \ensuremath{\D{P}} for the element at position \ensuremath{\V{v}\;\mathbin{:}\;\D{Fin}\;\V{n}} in a \ensuremath{\D{Vec}}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[4]{}\F{lookup}\;\mathbin{:}\;\{\mskip1.5mu \V{xs}\;\mathbin{:}\;\D{Vec}\;\V{A}\;\V{n}\mskip1.5mu\}\;\to \;\D{Fin}\;\V{n}\;\to \;\D{All}\;\D{P}\;\V{xs}\;\to \;\D{P}\;\V{x}{}\<[E]%
\\
\>[4]{}\F{lookup}\;\Con{zero}\;(\V{px}\;\Con{::}\;\anonymous )\;\mathrel{=}\;\V{px}{}\<[E]%
\\
\>[4]{}\F{lookup}\;(\Con{suc}\;\V{idx})\;(\anonymous \;\Con{::}\;\V{pxs})\;\mathrel{=}\;\F{lookup}\;\V{idx}\;\V{pxs}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

An important application of dependent types is to encode programming languages
syntax. The role of dependent types in this domain is to encode programs that
only allow well-typed and well-scoped terms~\cite{Benton2012}. Intuitively, we encode
the static semantics of the object language in the host language AST's
constructor, leaving the responsibility of checking type safety to the
host's language type-checker. Before considering our solution to encode list-machine
programs as instrisincally-typed syntax, we review this approach in a reduced example.

Consider the following simple
expression language. For this paper, we assume a basic knowledge of functional
programming and Agda. %\footnote{For further information about Agda, see~\cite{Norell2009,Stump16}.}.

@@

With this data type, we can construct expressions
to denote terms that should not be considered well-typed like
\ensuremath{(\Con{Num}\;\V{1})\;\F{+}\;\Con{True}}. Using this approach, we need to specify the static semantics
as another definition, which should consider all possible cases to avoid the
definition of ill-typed terms.

@@

@@

In this section, we review the syntax, semantics and typing rules for the
list-machine, as originally proposed by Appel et. al.~\cite{AppelDL12}.
The machine is a simple pointer virtual machine where values
are empty lists and cons-cells:
\[
a ::= nil\,\mid\,cons(a_1,a_2)
\]
Throughout this text, the meta-variable $a$ is used to denote an arbitrary value, $v$ denotes a variable
and $l$ a program label. Following common practice, all meta-variables can
appear primed or subscripted. The syntax of the virtual machine instructions are
presented next.
\begin{figure}[H]
\[
\begin{array}{rcll}
  \iota & ::=  & \text{jump }l                       & \text{(jump instruction)}\\
        & \mid & \text{branch-if-nil $v$~$l$}            & \text{(if $v = nil$ goto $l$)}\\
        & \mid & \text{fetch-field $v$ 0 $v'$}       & \text{(fetch the head of $v$ into $v'$)}\\
        & \mid & \text{fetch-field  $v$ 1 $v'$}      & \text{(fetch the tail of $v$ into $v'$)}\\
        & \mid & \text{cons $v_0$ $v_1$ $v'$}        & \text{(make a cons cell in $v'$)} \\
        & \mid & \text{halt}                         & \text{(finishes execution)}\\
        & \mid & \iota_1;\iota_2                             & \text{(sequential composition)}\\
      p & ::=  & l_i \,:\,\iota\,;\,p                    & \text{(program: sequence of blocks)}\\
        & \mid & \text{end}                          & \text{(end of block list)}\\
\end{array}
\]
\centering
\caption{Syntax of list-machine programs}
\label{fig:list-syntax}
\end{figure}
A program is just a sequence of blocks referenced by a unique label.

Each program variable is assigned to a list type (Figure~\ref{fig:type-syntax}), which is used to guarantee the safety of executing
fetch-field instructions that demands non-empty list arguments. In order to express such refinements, types are subject to a
subtyping relation. The meta-variable $\tau$ denotes an arbitrary type.
\begin{figure}[H]
\[
\begin{array}{rcll}
  \tau & ::=  & \text{nil} & \text{(type for empty lists)}\\
       & \mid & \text{list }\tau & \text{(lists whose elements have type $\tau$)}\\
       & \mid & \text{listcons }\tau & \text{(non-empty lists of $\tau$)}\\
\end{array}
\]
\centering
\caption{Type syntax}
\label{fig:type-syntax}
\end{figure}
The notation $\tau \subset \tau'$
denotes the subtyping judgment, which is defined as follows.
\begin{figure}[H]
\[
\begin{array}{c}
  \inference{}
            {\tau\subset \tau}
            [subtype-refl]
  \\ \\
  \inference{}
            {nil \subset list\:\tau}
            [subtype-nil]
  \\ \\
  \inference{\tau \subset \tau'}
            {list\:\tau\subset list\:\tau'}
            [subtype-list]\\ \\
  \multicolumn{3}{c}{
  \inference{\tau \subset \tau'}
            {listcons\:\tau\subset list\:\tau'}
            [subtype-listmixed]} \\ \\
  \multicolumn{3}{c}{
            \inference{\tau \subset \tau'}
            {listcons\:\tau\subset listcons\:\tau'}
            [subtype-listcons]} \\ \\
\end{array}
\]
\centering
\caption{Subtyping relation}
\label{fig:subtyping}
\end{figure}
Basically, the subtyping relation specifies that $nil$ (empty list type) is
subtype of any list type and $listcons\:\tau$ is a subtype of the $list\:\tau$.
The other rules specify that type constructors $list$ and $listcons$ respect
the subtyping relation. The least common supertype $\tau = \tau_1 \sqcap \tau_2$ of
$\tau_1$ and $\tau_2$ is defined as the smallest type such that $\tau_1$ and $\tau_2$
are subtypes of $\tau$.

Following the common practice, the meta-variable $\Gamma$ denotes an
environment binding variable names to their types. Notation $\{\}$ denotes an empty environment, $v : \tau , \Gamma$
denotes the operation of including a new entry for variable $v$ with type $\tau$
in $\Gamma$ and $\Gamma [v := \tau]$ denotes the environment which is identical to $\Gamma$, except
by the entry which associates variable $v$ with type $\tau$.
Subtyping is extended to contexts as follows.
\begin{figure}[H]
\[
\begin{array}{cc}
  \inference{}
            {\Gamma \subset_{env} \{\}}[b1]
  &
  \inference{\tau' \subset \tau & \Gamma' \subset_{env} \Gamma_2}
            {v : \tau' , \Gamma' \subset_{env} v : \tau , \Gamma_2}
            [b2]
\end{array}
\]
\centering
\caption{Subtyping relation for typing contexts.}
\label{fig:subtyping-context}
\end{figure}
The variable $\Pi$ is used to denote \emph{program typings}, i.e. finite mappings between
labels and typing contexts $\Gamma$, where notation $\Pi(l) = \Gamma$ denotes that
$\Gamma$ stores the types of variables on the entry point of the block labeled by $l$.
Using the previously defined notations, the typing rules for the list-machine are defined
as a syntax-directed judgment $\Pi \vdash_{\text{instr}} \Gamma \{ \iota \} \Gamma'$,
which intuitively means that the instruction $\iota$ transforms an input typing environment
$\Gamma$ into an output environment $\Gamma'$ under a fixed program typing $\Pi$. The typing
rules for the list-machine instructions are defined as follows.
